---
title: C Book Club for R Contributors
subtitle: "Session 1: Modern C Chapters 1-3"
author: Heather Turner
date: last-modified
date-format: long
format: 
  warwickpres-revealjs: default
---

# Preface {.inverse}

## Getting started

- IDE
   - RStudio + Terminal
   - VS Code + C extension / VSCodium (FOSS version)
   - Emacs 
   - Others: CLion ($), Vim, XCode (macOS) ..., or text editor + terminal
 - Compiler, e.g. c99, gcc and clang
   - clang and gcc give better diagnostics
   - Installing XCode on macOS / RTools on Windows gives you clang / gcc
 - Modern C example code download has moved: <https://inria.hal.science/hal-03345464>

# Level 0:<br> ENCOUNTER {.inverse background-image="media/magpie.jpg"}
 
# Chapter 1: Getting Started {.inverse}

## getting-started.c

```c
/* This may look like nonsense, but really is -*- mode: C -*- */
#include <stdlib.h>
#include <stdio.h>

/* The main thing that this program does. */
int main(void) {
  // Declarations
  double A[5] = {
    [0] = 9.0,
    [1] = 2.9,
    [4] = 3.E+25,
    [3] = .00007,
  };

  // Doing some work
  for (size_t i = 0; i < 5; ++i) {
     printf("element %zu is %g, \tits square is %g\n",
            i,
            A[i],
            A[i]*A[i]);
  } 

  return EXIT_SUCCESS;
}
```

## Takeaways

- C is an imperative programming language.
    - The program is a set of orders to the computer to perform tasks
- C is a compiled programming language
    - The compiler translates code info the binary code or executable
    - The target code is platform-dependent
        - Running and testing a program on a single platform does not guarantee portability
        
## Exs 3 and 5: Compile then execute

Find which compilation command works for you.

 - -Wall = warn all
 - -o: where to store compiler output
 - -lm tells it to add some standard mathematical functions if necessary

```c
$ clang -Wall -lm -o getting-started getting-started.c
$ ./getting-started
```
```
element 0 is 9,         its square is 81
element 1 is 2.9,       its square is 8.41
element 2 is 0,         its square is 0
element 3 is 7e-05,     its square is 4.9e-09
element 4 is 3e+25,     its square is 9e+50
```

## Exs 6: Fix a bad program

```c
clang -Wall -o bad bad.c
```
```
bad.c:4:1: warning: return type of 'main' is not 'int' [-Wmain-return-type]
void main() {
^
bad.c:4:1: note: change return type to 'int'
void main() {
^~~~
int
bad.c:16:6: error: call to undeclared library function 'printf' with type 'int (const char *, ...)'; ISO C99 and later do not support implicit function declarations [-Wimplicit-function-declaration]
     printf("element %d is %g, \tits square is %g\n", /*@\label{printf-start-badly}*/
     ^
bad.c:16:6: note: include the header <stdio.h> or explicitly provide a declaration for 'printf'
bad.c:22:3: error: void function 'main' should not return a value [-Wreturn-type]
  return 0;                                           /*@\label{main-return-badly}*/
  ^      ~
```

## Exs 6 (ctd): fix bad.c

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2|5"
/* This may look like nonsense, but really is -*- mode: C -*- */
#include <stdio.h>

/* The main thing that this program does. */
int main() {
  // Declarations
  int i;
  double A[5] = {
    9.0,
    2.9,
    3.E+25,
    00007
  };

  // Doing some work
  for (i = 0; i < 5; ++i) {
    
    printf("element %d is %g, \tits square is %g\n",
           i,
           A[i],
           A[i]*A[i]);                           
  }
  
  return 0;                                           
}
```

## Exs 7: correct array definition

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "9-12"
/* This may look like nonsense, but really is -*- mode: C -*- */
#include <stdio.h>

/* The main thing that this program does. */
int main() {
  // Declarations
  int i;
  double A[5] = {
    [0] = 9.0,
    [1] = 2.9,
    [4] = 3.E+25,
    [3] = .00007
  };

  // Doing some work
  for (i = 0; i < 5; ++i) {
    
    printf("element %d is %g, \tits square is %g\n",
           i,
           A[i],
           A[i]*A[i]);                           
  }
  
  return 0;                                           
}
```

## Compilation

- A C program should compile cleanly without warnings.
- Use `-Werror` to not create an executable even when there are just warnings

# Chapter 2: The principal structure of a program {.inverse}

## Grammar (1)

 - special words: `#include`, `int`, `void`, `double`, `for`, `return`
     - cannot be changed
 - brackets: `{}`, `()`, `[]`, `/**/`, and `<>`
 - comma: separate arguments or list elements 
     - optionally after last element in list
 - semicolon: terminates a statement inside `{}`
 - comments: `/* */` multiline, `//` one line
 - literals: fixed values that are part of the program
     - use double quotes for string literals
     
## Grammar (2)

 - identifiers: names that we give to things 
     - data objects/variables e.g. `A`, `i`
     - type aliases, e.g. `size_t`, specify sort of object
         - _t reminds us the identifier is a type
 - functions e.g . `main`
     - `main` must be present, it is the starting point of execution
 - constants e.g. `EXIT_SUCCESS`
 - operators, like R except:
     - `=` for initialization and assignment
     - `++` to increment

## Exs 9: brackets

 - `<>` only for specifying headers to include
 - `{}` embracing body of function, embracing elements of array (vector),
embracing multi-line conditional statement
 - `[]` specifying length of array, specifying elements of array, indexing
 
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2-3|6,8,13,16,21,24|8-13,19-20"
/* This may look like nonsense, but really is -*- mode: C -*- */
#include <stdlib.h>
#include <stdio.h>

/* The main thing that this program does. */
int main(void) {
  // Declarations
  double A[5] = {
    [0] = 9.0,
    [1] = 2.9,
    [4] = 3.E+25,
    [3] = .00007,
  };

  // Doing some work
  for (size_t i = 0; i < 5; ++i) {
     printf("element %zu is %g, \tits square is %g\n",
            i,
            A[i],
            A[i]*A[i]);
  } 

  return EXIT_SUCCESS;
}
```

## Declarations

All identifiers have to be declared (maybe not directly in our program)

 - `int main(void)`
    - a function of type int
 - `double A[5]`
    - an array of items of type double, of length 5
 - `size_t i` 
    - i is of type size_t
 - `printf` declared in `stdio.h`
 - `size_t` and `EXIT_SUCCESS` come from `stdlib.h`
 
   ```c
   int printf(char const format[static 1], ...); 
   typedef unsigned long size_t;
   #define EXIT_SUCCESS 0
   ```
     
To look up something declared in a header file
```sh
apropos printf #loads of stuff
man printf #only printf R-like help file
man 3 printf #other related functions e.g. fprintf
```

## Identifiers

 - Declaration limited to scope
    - block scope: limited within enclosing `{}`, e.g. for loop
    - file scope: not within `{}`, so limited by file -> globals
 - Definition: specifies object itself
    - simplification: always initialize variables
        - augment declaration with initial value
        - use `[]` to designate specific elements (e.g. in array)
        - missing elements default to 0
        - indices from zero: think of as distance from start of array
    - can have many declarations, but need exactly one definition
    
## Statements (1)

Instructions telling compiler what to do with declared identifiers

 - Iteration, e.g. `for` 
    - loop body between `{}`
    - arguments separated by semicolons
        1. declaration, definition and initialization of loop variable
        2. loop condition
        3. statement executed after each iteration: ++i
    - *don't* define loop variable outside `()` or reuse for several loops
 - Function calls
    - C uses "call by value": values of identifiers cannot be changed by fn
    
## Statements  (2)

 - Function return
    - last statement in function
    - must return value of declared type
    - brackets not required around return value
  
         
:::speaker-notes
 (vs "call be reference" which allows value of variable to be changed)
:::

# Level 1:<br> ACQUAINTANCE {.inverse background-image="media/raven.jpg"}

## Buckle Up

Details about how objects are defined. Some main points

- focus on unsigned integers
- gradually introduce pointers
- use arrays where possible

## Indentation Style

Recommends ["One True Brace Style"](https://en.wikipedia.org/wiki/Indentation_style#Variant:_1TBS_(OTBS))

```c
int main (int argc , char * argv [ argc +1]) {
  puts (" Hello   world !");
  if ( argc > 1) {
    while ( true ) {
      puts (" some   programs   never   stop ");
    }
  } else {
    do {
      puts ("but   this  one   does ");
    } while ( false );
  }
  return EXIT_SUCCESS ;
}
```

There is no standard in base R (see R Dev Guide  [#133](https://github.com/r-devel/rdevguide/issues/133)) but K&R is used more and just a minor variant (as above, but first `{` on new line).

# Chapter 3. Everything is about control {.inverse}

## If else blocks

As in R

 - 0 represents logical false
 - Any value different from 0 represents logical true.

- Don’t compare to 0, false, or true as redundant
    - `if(i)` [found 5 times in base R](https://github.com/search?q=repo%3Ar-devel%2Fr-svn+%22if+%28i%29%22+path%3A*.c&type=code)
    - `if (i != 0)` [found once](https://github.com/search?q=repo%3Ar-devel%2Fr-svn+%22if+%28i+%21%3D+0%29%22+path%3A*.c&type=code)
    
## Exs 1: Add in for loop 

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "18-23"
/* This may look like nonsense, but really is -*- mode: C -*- */
#include <stdlib.h>
#include <stdio.h>
  
/* The main thing that this program does. */
int main(void) {

  // Declarations
  double A[5] = {
    [0] = 9.0,
    [1] = 2.9,
    [4] = 3.E+25,
    [3] = .00007,
  };

  // Doing some work
  for (size_t i = 0; i < 5; ++i) {
    if (i) {
      printf("element %zu is %g, \tits square is %g\n",
             i,
             A[i],
             A[i]*A[i]);   
    }
  }
  
  return EXIT_SUCCESS;
}
```

## Exs 1: Add in for loop - result

```
$ ./getting-started-if
element 1 is 2.9,       its square is 8.41
element 2 is 0,         its square is 0
element 3 is 7e-05,     its square is 4.9e-09
element 4 is 3e+25,     its square is 9e+50
```

## `for` loops

We can have multiple loops iterating on `i` as long as scopes don't overlap. [?]

[Exs 2] Try to imagine what happens when `i` has value 0 and is decremented by means of the operator `--`

 - Becomes null?
 
## `while` and `do`

`while` works as in R

`do` is similar to `while` but the condition after the dependent block:

```c
do { // Iterates
  x *= (2.0 - a*x); // Heron approximation
} while ( fabs (1.0 - a*x) >= eps ); // Iterates until close
```

=> if condition false, will iterate once before closing.

Using `{}` recommended. Note `;` after while statement after `do {}`.

## `break` and `continue`

`break` works as in R

`continue` is equivalent to `next` in R

`for (;;)` is equivalent to `while(true)` (former appears 23 times; latter just once in R sources)

## heron.c

```c
#include <stdlib.h>
#include <stdio.h>

/* lower and upper iteration limits centered around 1.0 */
static double const eps1m01 = 1.0 - 0x1P-01;
static double const eps1p01 = 1.0 + 0x1P-01;
static double const eps1m24 = 1.0 - 0x1P-24;
static double const eps1p24 = 1.0 + 0x1P-24;

int main(int argc, char* argv[argc+1]) {
  for (int i = 1; i < argc; ++i) {        // process args
    double const a = strtod(argv[i], 0);  // arg -> double
    double x = 1.0;
    for (;;) {                    // by powers of 2
      double prod = a*x;
      if (prod < eps1m01) {
        x *= 2.0;
      } else if   (eps1p01 < prod) {
        x *= 0.5;
      } else {
        break;
      }
    }
    for (;;) {                    // Heron approximation
      double prod = a*x;
      if ((prod < eps1m24) || (eps1p24 < prod)) {
        x *= (2.0 - prod);
      } else {
        break;
      }
    }
    printf("heron: a=%.5e,\tx=%.5e,\ta*x=%.12f\n",
           a, x, a*x);
  }
  return EXIT_SUCCESS;
}
```

## Exs 4: Analyze `heron.c` using `printf`

Add condition print statement to loops, e.g.
```c
if (i == 1) printf("i = %d, loop 1, x: %f\n", i, x);
```

```
$./heron 0.07 5 6E+23
i = 1, loop 1, x: 2.000000
i = 1, loop 1, x: 4.000000
i = 1, loop 1, x: 8.000000
i = 1, loop 1, x: 16.000000
i = 1, loop 2, x: 14.080000
i = 1, loop 2, x: 14.282752
i = 1, loop 2, x: 14.285714
heron: a=7.00000e-02,   x=1.42857e+01,  a*x=0.999999957002
```

## Exs 5: Describe the use of the parameters `argc` and `argv`

```c
int main(int argc, char* argv[argc+1]) {
  printf("argc: %d\n", argc);
  printf("argv[0]: %s\n", argv[0]);
...  
```

```
$./heron 0.07 5 6E+23
argc: 4
argv[0]: ./heron
argv[1]: 0.07
argv[2]: 5
argv[3]: 6E+23
argv[4]: (null)
```

`main` receives an array of pointers to char: the program name, argc-1 program arguments, and one null pointer that terminates the array.

## Exs 6: Print and change `eps1m01`

Add condition print statement to loops, e.g.
```c
  printf("eps1m01: %f\n", eps1m01);
```

```
$./heron 0.07 5 6E+23
eps1m01: 0.750000
i = 1, loop 1, x: 2.000000
i = 1, loop 1, x: 4.000000
i = 1, loop 1, x: 8.000000
i = 1, loop 1, x: 16.000000
i = 1, loop 2, x: 14.080000
i = 1, loop 2, x: 14.282752
i = 1, loop 2, x: 14.285714
heron: a=7.00000e-02,   x=1.42857e+01,  a*x=0.999999957002
```

Originally `eps1m01: 0.5`.

## `switch`

In R, `switch` only returns output for one matching case.

In C, `switch` jumps to matching case and continues to end or break.

## Exs 7: Leaving out break statement 

With help from  <https://overiq.com/c-programming-101/the-switch-statement-in-c/> to get character input from `scanf`

```
#include <stdlib.h>
#include <stdio.h>

int main(void){
  char arg;
  
  printf("Enter arg: ");
  scanf(" %c", &arg);
  
  switch (arg) {
  case 'm': puts (" this  is a  magpie ");
    break ;
  case 'r': puts (" this  is a  raven ");
    break ;
  case 'j': puts (" this  is a  jay ");
   // break ;
  case 'c': puts (" this  is a  chough ");
    break ;
  default : puts (" this  is an  unknown   corvid ");
  }
  
  return EXIT_SUCCESS;
}
```

## Exs 7: output

```
$ ./switch
Enter arg: r
 this  is a  raven 
$ ./switch
Enter arg: j
 this  is a  jay 
 this  is a  chough 
```